Propositions as Sessions, Mechanically

### 2015-04-03 paper/formalisation.tex
Updated paper: CP section and added figures.

The formalisation was performed using the Coq proof
assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
challenge~\cite{Aydemir:2005:MMM}, the development utilises available
libraries and infrastructure as much as possible. In particular,
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
library for handling, among other things, association lists (environments) and
free variable representation. The current development uses the version of the
library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
has been updated for Coq version 8.4pl4. One goal of the formalisation effort
was to minimise the extent to which the development differs from the
pen-and-paper presentation and avoid questions of adequacy of the
representation. Some changes were made to ease the mechanisation and are
highlighted where they occur.

@@

%\input{with-connect-rule}
\input{cp}

The locally nameless first-order representation for binders and free variables
was adopted for both GV and CP, following
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
variant prevents variable-capture issues when compared with other first-order
approaches. Alternatively, I could have chosen a higher-order abstract syntax
(HOAS) approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, it is less clear how to handle linearity using this
approach and has the disadvantage of being able to represent terms not valid
in the language being mechanised (so called ``exotic'' terms). Further, a
higher-order approach would prevent the use of the Metatheory library which
provides a number of useful tactics for manipulating typing contexts.

@@

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
set of variables from being considered as the binder. In contrast, the
traditional ``exists-fresh'' approach, where the binder is only required to be
fresh within the abstraction's body, does not produce strong enough induction
principles in some cases. This change will necessitate changes to the typing
rules of GV and CP. Consider for instance the Connect rule in GV (C) and its
corresponding cofinite quantification version (CCO) in
Figure~\ref{fig:connect}. The conclusion specifies a variable in (C) but a
type in (CCO) (de Bruijn index rather than a name). Thus, in the premise for
(CCO) one can choose a suitably fresh $x$ in each premise. Note that, $L$ can
be chosen such that $\Psi \cup \Phi \cup \fv(M) \cup \fv(N) \subseteq L$,
where $\fv(M)$ are the set of free variables in term $M$, and $x$ can be the
same in both premises. Similar changes are made to the rules for CP. These
changes do not alter the typing of processes nor the semantics of the
translation from GV to CP. An equivalence between ``exists-fresh'' and
cofinite quantification definitions of the simply-typed $\lambda$-calculus is
given by \citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect
a similar result could be derived straightforwardly (but tediously) for the
CP/GV system.

Another aspect of the encoding is how to define terms. The intrinsic encoding
described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
approach, indexing terms by their type so as to prevent ill-formed terms from
being constructed. However, in the encoding for CP/GV there are issues with
using this encoding: $(1)$ it is not immediately clear how to handle linear
contexts using the de Bruijn variable encoding, since in the intrinsic setting
the environment automatically supports weakening; $(2)$ an intrinsic encoding
of GV terms would be complicated by the need to enforce that session types
occur in certain instances (by use of a predicate); and $(3)$ well-typed terms
require extra assumptions about binder freshness which cannot reasonably be
expressed as a function type. For these reasons, an intrinsic approach does
not offer much benefit since a well-typed term relation would still need to be
defined. Therefore, I chose an extrinsic encoding for terms and define
well-typed terms as a separate inductive type.

@@

\citeauthor{Park:2014:MMW} describe a technique for
removing nonlinear contexts from typing judgements. While their work extended
to \fpop it is not clear how to handle an environment containing both
non-linear and linear types as in GV. I wish to maintain as close a
relationship as possible to the paper system presented by
\citeauthor{Wadler:2014}, so separating out the non-linear and linear
components (as in \fpop) is not an option at this stage.

@@

@@

Formalising the process calculus CP provided many challenges for the
metatheory library since to my knowledge, the library has never been applied
to a process calculus before. So for instance, in the lambda calculus one
typically does not work up to permutation of binders, whereas in a process
calculus these kinds of structural equivalences are standard. Additionally,
substitution and opening are only defined on processes for names. In other
words, one cannot substitute an arbitrary process term for a name, unlike in
the lambda calculus. Thus frameworks which assume a lambda calculi style
substitution operation are not immediately amenable to the CP
setting~\cite{Lee:2012}.

Note that the rules for polymorphism, $\exists$ and $\forall$, are defined but
cut elimination on them is not. Since GV (\S~\ref{sec:gv}) does not support
polymorphism the reduction rules are unnecessary for the translation. Future
work is to support reduction under these terms.

Initially, an implicit notion of weakening and contraction were chosen so as
to be in line with \citeauthor{Wadler:2014}'s definitions. However, explicit
weakening and contraction were required to differentiate the reduction rules
for the various interactions between servers and clients.

Ordering in environments is ignored and the development required lemmas
allowing the permutation of environments. Support for permutating environments
is not currently supported in the Metatheory library but some general lemmas
were added as a result of this work. Additionally, custom tactics were
developed to automatatically solve some simple permutation goals. However,
these tactics are not a ``catch all'' and some generalising still needs to be
done.

The paper presentation omits details: the principal cut rules do not specify
weakening could occur: since weakening is implicit is tacitly assumed
weakening can always be pushed further up the derivation tree e.g. AxCut rule

another interesting aspect which the formalism reveal is the rule for
associativity is constructed in a way to permit associativity in a certain way
pre-determined

@@

@@

@@

>>>>

The formalisation was performed using the Coq proof
assisstant~\cite{Coq:manual}. In the spirit of the POPLMARK
challenge~\cite{Aydemir:2005:MMM}, the development utilises available
libraries and infrastructure as much as possible. In particular,
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM} developed a metatheory
library for handling, among other things, association lists (environments) and
variable representation. The current development uses the version of the
library distributed by~\citeauthor{Park:2014:MMW}~\cite{Park:2014:MMW} which
has been updated for Coq version 8.4pl4. One goal of the formalisation effort
was to minimise the extent to which the development differs from the
pen-and-paper presentation and avoid questions of adequacy of the
representation. Some changes were made to ease the mechanisation and are
highlighted where they occur.

@@

The locally nameless cofinite quantification (LNCQ) representation for binders
and free variables was adopted for both GV and CP, following
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}. This first-order
variant prevents variable-capture issues when compared with other first-order
approaches. Alternatively, I could have chosen a higher-order abstract syntax
(HOAS) approach such as the parametric version (PHOAS) described
by~\citeauthor{Chlipala:2008:PHOAS}~\cite{Chlipala:2008:PHOAS}. PHOAS relies
on the metalanguage (Coq, in this case) for handling binders and substitution
so has less overhead in managing syntax than the locally nameless
approach. However, it is less clear how to handle linearity using this
approach and has the disadvantage of being able to represent terms not valid
in the language being mechanised (so called ``exotic'' terms). Further, a
higher-order approach would prevent the use of the Metatheory library which
provides a number of useful tactics for manipulating typing contexts.

@@

For handling binder freshness, I chose to adopt the cofinite quantification
approach described by~\citeauthor{Aydemir:2008:EFM}, which excludes a finite
set of variables from being considered as the binder. In contrast, the
traditional ``exists-fresh'' approach, where the binder is only required to be
fresh within the abstraction's body, does not produce strong enough induction
principles in some cases, e.g. weakening~\cite{Aydemir:2008:EFM}. This
decision impacts upon the premises of the CP/GV typing rules (\S~\ref{sec:cp}
and \S~\ref{sec:gv}), however these changes do not alter the typing of
processes nor the semantics of the translation from GV to CP. An equivalence
between ``exists-fresh'' and cofinite quantification definitions of the
simply-typed $\lambda$-calculus is given by
\citeauthor{Aydemir:2008:EFM}~\cite{Aydemir:2008:EFM}, and I suspect a similar
result could be derived straightforwardly (but tediously) for the CP/GV
system.

Another aspect of the encoding is how to define terms. The intrinsic encoding
described by~\citeauthor{Benton:2012:STT}~\cite{Benton:2012:STT} is one
approach, indexing terms by their type so as to prevent ill-formed terms from
being constructed. However, there are issues with using this encoding for
CP/GV: $(1)$ it is not immediately clear how to handle linear contexts using
the de Bruijn variable encoding, since in the intrinsic setting the
environment automatically supports weakening; $(2)$ an intrinsic encoding of
GV terms would be complicated by the need to enforce that session types occur
in certain instances (by use of a predicate); and $(3)$ well-typed terms
require extra assumptions about binder freshness which cannot reasonably be
expressed as a function type. For these reasons, an intrinsic approach does
not offer much benefit since a well-typed term relation would still need to be
defined. Therefore, I chose an extrinsic encoding for terms and define
well-typed terms as a separate inductive type.

@@

\citeauthor{Park:2014:MMW} describe a technique for removing non-linear
contexts from typing judgements. While their work extended to \fpop it is not
clear how to handle an environment containing both non-linear and linear types
as in CP/GV. I wish to maintain as close a relationship as possible to the
pen-and-paper system presented by \citeauthor{Wadler:2014}, so separating out
the non-linear and linear components (as in \fpop) is not an option at this
stage.

@@

\input{cp}

@@

Formalising the process calculus CP provided many challenges for the
metatheory library since to my knowledge, the library has never been applied
to a process calculus before. So for instance, in the lambda calculus one
typically does not work up to permutation of binders, whereas in a process
calculus these kinds of structural equivalences are standard. Additionally,
substitution and opening are only defined for names. In other words, one
cannot substitute an arbitrary process term for a name, unlike in the lambda
calculus. Thus frameworks which assume a lambda calculi style operations are
not immediately amenable to the CP setting~\cite{Lee:2012}. Ordering in
environments is ignored so the development required lemmas allowing the
permutation of environments. Support for permutating environments is not
currently supported by the Metatheory library but some general lemmas were
added as a result of this work. Additionally, custom tactics were developed to
automatatically solve some simple permutation goals, however these tactics are
not a ``catch all'' and some generalising still needs to be done.

% Description of the typing rules...

Figure~\ref{fig:cll} presents the process calculus CP in LNCQ form (ignoring
permutation premises). In other words bound variables are represented using de
Brujin indices and free variables are represented using an abstract type
provided by the Metatheory library. The rules with binders reflect this by
specifying the type of a bound variable in the constructor rather than the
name. Cofinite quantification is performed in the hypotheses to these rules by
excluding a finite set of names from opening the subprocesses. Note that the
rules for polymorphism, $\exists$ and $\forall$, were included in the final
Coq development since GV (\S~\ref{sec:gv}) does not support polymorphism
making the rules unnecessary for the translation. They appear in the Figure
simply for the sake of completeness. Additionally, as noted by
\citeauthor{Lee:2012}, support for polymorphism in type variables creates an
explosion of the infrastructure requiring duplication of opening and
substituting definitions and lemmas.

% ...leading on to the discussion of the explicit weakening...

Initially, an implicit notion of weakening was chosen so as to be in line with
\citeauthor{Wadler:2014}'s definitions. However, explicit weakening is
required to differentiate the principal cut elimination rules for the various
interactions between servers and clients. contraction is not supported because
of the difficulty in specifying its principal cut elimination rule. In order
to specify the rule one requires to show the following:

\todo{v A -> !0(A).P | Q{x/x'}
  ==>
v A -> !0(A).P | v A -> !0(A).P | contract (Gamma) Q}

Unfortunately, the reduction relation does not have knowledge of the
environment typing process P. This fact does not impact upon expressing
weakening since for weakening only the names (not the bindings) within the
environment are required. The paper presentation does not specify where
weakening could occur; it is tacitly assumed weakening can always be pushed
further up the derivation tree e.g. AxCut rule. a consequence of this choice
is that for the empty choice rule there may only be a single binding in the
context (cf. an arbitrary context in \citeauthor{Wadler:2014}'s
presentation~\cite{Wadler:2014}).

% ...leading on to presentation of principal cut elimination rules...

\input{principalcp}

The principal cut elimination rules are presented in Figure~\ref{??}. These
rules are expressed to indicate some additional complexity not present in the
original presentation. For example, structural equivalences and associativity
are not unique and different process terms result depending on initial binder
permutation. ${0 <~> 1}P$ indicates the permuting of binders $0$ and $1$ in
process P. The paper presentation specifies only one possible associativity
rule lending to the specific split of the linear environments. The reduction
rules were complicated to prove due to being composed of multiple constructor
instances and differed significantly to the standard operational semantics one
typically deals with in lambda calculus formalisms.

% ...commuting conversions

\input{commconvonecp}

\input{commconvtwocp}

% ...leading on to a discussion of further CP-specific issues encountered.

@@

\input{gv}

@@

another issue is the way i tried to formulate types initially treating them as
indexed families indexed by kinds. while this exploits the dependently typed
approach it complicated later definitions and proofs. for instance decidablity
of equality on types now relied on an axiom permitting heterogeneous
equality. further when formulating the typing environment we must decide which
kind each type had and most rules are polymorphic in that decision leading to
the creation of an awkward inspection function. these issues highlighted that
one should be careful and considerate how the representation will affect later
expressiveness.

explicit limiting and unlimiting of abstraction to ensure rules are syntax
directed - permits inductive argument on terms when translating from gv to cp

@@

type inference of gv terms required cluttered definition of translation with
impossible error cases. avoided a dependently typed solution due to
difficulties encountered earlier in the development. james mckinna et
al. suggest encoding the successful cases in an inductive type and then
defining the function to act on these cases. will this really work? this means
a graph of positive cases for types then a function definition for it too

translation uses the coq-ext-lib from michela for monadic notation
